Complexity-stability relationships in large, complex systems

### 2023-03-21 papers/general-stability.tex
update draft

\usepackage{colortbl}
\usepackage{graphicx} 
\usepackage{booktabs}
\usepackage{caption}
\newcommand{\red}{\textcolor{red}}

@@

\begin{abstract}
Fifty years ago, Robert May predicted that large, complex systems are unlikely to be stable. Here, we revisit May's argument to show that there are, in fact, two kinds of complexity-stability relationships in disordered dynamical systems, depending on the relative convexity of the response to self- vs. cross-interactions. We illustrate the transition between "May" (complexity begets instability) and "anti-May" (complexity begets stability) behavior with a non-linear generalization of the Lotka-Volterra model. 
\end{abstract}

@@

Few mathematical arguments have influenced biological thinking like May's prediction that large, complex systems cannot be stable. It is, on the face of it, a perplexing conclusion. On the one hand, May's argument is so simple and compelling that it hard not to conclude that it must hold universally. On the other, it is clear that at least some large, complex systems are stable---or else which stable patterns would biology be studying in the first place? If anything, the opposite relationship between complexity and stability seems to hold empirically: rich, strongly coupled ecosystems (eg. rainforests) tend to be  stable over time, while sparser ones (eg. arctic environments) often exhibit large fluctuations, exctinctions, and invasions.

May's argument goes as follows. A system with $N$ populations $x_i$ is stable is one that can be characterized by an equilibrium point $x^*$. Near that equilibrium $x = x^* + \delta x$, the dynamics of the system will be described by linear equations $d(\delta x)/dt = A (\delta x)$, and stability of these equations requires that all eigenvalues of $A$ have negative real part. But if $A$ can be represented as $A = B - I$, where $B$ consists of random, independent interactions with zero mean and variance $\sigma^2$, and $-I$ corresponds to stabilizing self-interactions on some natural timescale, then the circular law of random matrix theory implies that all eigenvalues of $A$ will have negative real part only if $\sigma^2 N < 1$. This places a sharp constraint on both diversity $N$ and interaction strengh $\sigma$, often referred to as "complexity begets instability". (This argument generalizes to $\langle B_{ij}\rangle \neq 0$, to incomplete connectivity, or to correlated interactions.)

In ecology, many authors have sought to ease the tension between May's prediction and empirical observation by invoking effects not captured by dynamical systems with random coefficients, such as spatial or food-web structure, time delays, etc. 
Here, we show that May's argument is itself incomplete: random dynamical systems do not necessarily imply that stability decreases with diversity or interaction strength---the opposite behavior is also possible, without the need for special or additional structure.

@@

Consider the dynamical system in $N$ variables
\begin{equation}\label{dynamics}
    \dot{x}_i = f(x_i) + \sum_{j}a_{ij}g(x_i)h(x_j) \, .
\end{equation}
Here $f(x_i)$ represents the contribution of a population $i$ to its own growth, while $g(x_i)$ and $h(x_j)$ capture the interaction of $i$ with other populations. 
That interaction is weighted by a coefficient $a_{ij}$, such that $a_{ij} > 0$ (resp. $a_{ji} < 0$) implies a positive (resp. negative) effect of $j$ on the growth of $i$. 
In the following we refer to $f(x_i)$ as the ``production function'' and $g(x_i)$ as the ``response function''.

The classic generalized Lotka-Volterra (GLV) competition model corresponds to $f$, $g$, $h$ all linear. 
However, it is natural both physically and biologically to consider more general functions, including power laws $f(x)\sim x^\alpha$, $g(x)\sim x^\beta$, $h(x) \sim x^\gamma$. 
From a physical perspective, we can imagine populations $x_i$ forming three-dimensional clusters whose growth is limited to their two-dimensional surface, leading to a production function $f(x) \sim x^{2/3}$. 
Biologically, the growth of organisms (populations of cells) has long been known to scale like $f(x) \sim x^k$ with $k\simeq 3/4$, which can be understood in terms of hydrodynamic constraints on vascular and pulmonary networks. 
For reasons that are not currently understood, a similar pattern of growth appears to recur at the community level. 
It is also common to model predator-prey interactions with a square root laws ($g(x) \sim h(x) \sim x^{1/2}$), for example.

@@

Now we decompose the interaction matrix as
$a = \mu \mathbf{1} + (\mu_s-\mu)I + S$,
with $\mathbf{1}$ the rank-one matrix with all entries equal to $1$
and $S$ a random matrix as defined above.
We can limit ourselves to consider 
$a = (\mu_s-\mu)I + S$ by relying on 
rank-one perturbation theory, we would only neglect outliers
uninfluential for the linear stbaility properties of the system by doing so
(see Ref.~[Stone2018]).
The Jacobian in Eqs.~\eqref{eq: jac off-diag} and \eqref{eq: jac diag} assumes therefore the form $M + LJR$ with
\begin{align}
    M & = \textrm{diag}\left(f_*'(\mathbf x) -
    \frac{g_*'(\mathbf x)f_*(\mathbf x)}{g_*(\mathbf x)}
    +(\mu_s-\mu)g_*(\mathbf x)h_*'(\mathbf x)\right) \ , \\
    L &= \textrm{diag}(g_*(\mathbf x)) \ , \\
    R &= \textrm{diag}(h_*'(\mathbf x)) \ .
\end{align}
The domain of the eigenvalues of $J^*$, defined by
Eq.~\eqref{eq: domain}, 
first touches the right half-plane at $\zeta = 0$, 
hence linear stability requires   
\begin{equation}\label{eq: random-stability}
    \sum_{i=1}^N \cfrac{\left(\cfrac{g_*(x_i)h_*'(x_i)}{f_*(x_i)}\right)^2}{
        \Bigl |\left(\cfrac{f_*'(x_i)}{f_*(x_i)} -
        \cfrac{g_*'(x_i)}{g_*(x_i)}\right)
        +(\mu_s-\mu) \cfrac{g_*(x_i)h_*'(x_i)}{f_*(x_i)} \Bigl |^2}
    < \sigma^{-2}. 
\end{equation}
This expression can be considered the
random interaction case equivalent of
the expression in Eq.~\eqref{homogeneous-general} 
which appears again in the denominator of the summand.
It reduces indeed to the homogenous case for $\sigma\to 0$.
\red{Discuss better.}

@@

In the general case in which the dependence on the equilibria
$\mathbf{x}_*$ does not cancel out, is not possible to
obtain a clean complexity-stability relationship as before.
We can make progress in the large $N$ limit by
trasforming the sum in an integral 
$\sum_{i=1}^N\to N\int dx_*P(x_*)$, where $P(x_*)$ 
is the equilibrium distribution of the system
depending on the statistic of the interactions,
and write
\begin{equation}\label{eq: random-stability int}
    \int dx_* \cfrac{NP(x_*) \left(\cfrac{g_*(x)h_*'(x)}{f_*(x)}\right)^2}{
        \Bigl |\left(\cfrac{f_*'(x)}{f_*(x)} -
        \cfrac{g_*'(x)}{g_*(x)}\right)
        +(\mu_s-\mu) \cfrac{g_*(x)h_*'(x)}{f_*(x)} \Bigl |^2}
    < \sigma^{-2} \ .
\end{equation}
This expression allows us to obtain insights on the stability properties of
the system in the case in which we are able to compute or estimate 
the probability distribution function $P(x_*)$.

@@

The equilibrium probability distribution function for $x_*$, $P(x_*)$,
can be obtained as the pushforward of the distribition of $\xi$.
The change of measure
$\left|d \xi/d x_*\right|=|\alpha-\beta|x_*^{\alpha-\beta-1}/\sigma \sqrt{N\langle x_*^{2\gamma}\rangle}$,
leads to
\begin{equation}\label{eq: dist general}
    P(x_*)=\frac{|\alpha-\beta|x_*^{\alpha-\beta-1}}{\sqrt{2\pi\sigma^2 N\langle x_*^{2\gamma}\rangle}}
    \exp{\left\{-\frac{(x_*^{\alpha-\beta}-\mu N\langle x_*^{\gamma}\rangle)^2}{2\sigma^2N\langle x_*^{2\gamma}\rangle}\right\}} \, ,
\end{equation}
where the expectations have to be computed self-consistently, 
taking also in account normalization constraint.

Let us focus to the case $\alpha=1$, $\beta=1.5$ and $\gamma=1$, with
$a_s=0$ in order to leverage the equilibrium solution in Eq.~\eqref{eq: dist general}.
which is tested against simulations in Fig.~\ref{fig: cavity sol.}.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=.45\textwidth]{figs/cavity.pdf}
        \caption{...}
        \label{fig: cavity sol.}
    \end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=.45\textwidth]{figs/beta1_5-S50-N10.pdf}
    \caption{...}
    \label{fig: stability line + sims}
\end{figure}

>>>>

\usepackage{colortbl}
\newcommand{\red}{\textcolor{red}}

@@

\begin{abstract}
Fifty years ago, Robert May predicted that large, complex systems are unlikely to be stable. Here, we revisit May's argument to show that there are, in fact, two kinds of complexity-stability relationships in disordered dynamical systems: in some cases, strong interaction and large diversity are, indeed, destabilizing; in other cases, they are stabilizing. Which class a given system falls into depends on the relative convexity of the response to self- vs. cross-interactions. We illustrate the transition between "May" (complexity begets instability) and "anti-May" (complexity begets stability) behavior with a non-linear  generalization of the Lotka-Volterra model. 
\end{abstract}

@@

Few mathematical arguments have influenced biological thinking like May's 1970 prediction that large, complex systems cannot be stable. It is, indeed, a perplexing prediction. On the one hand, the argument is so simple and compelling that it is hard not to conclude that it must hold universally. On the other, it is clear that at least some large, complex systems are stable---or else which stable patterns would biology be studying in the first place? If anything, the opposite relationship between complexity and stability seems to hold empirically: rich, strongly coupled ecosystems (eg. rainforests) tend to be  stable over time, while sparser ones (eg. arctic environments) often exhibit large fluctuations, exctinctions, and invasions.

May's argument can be paraphrased as follows. If a system with $N$ populations $x_i$ is stable, then it can be characterized by an equilibrium $x^*$. Near that equilibrium $x = x^* + y$, the dynamics of the system will be described by linear equations $\delta\dot{y} = A y$, and stability of these equations requires that all eigenvalues of $A$ have negative real part. But if $A$ can be represented as $A = B - I$, where $B$ consists of random, independent interactions with zero mean and variance $\sigma^2$, and $-I$ corresponds to stabilizing self-interactions on some natural timescale, then the circular law of random matrix theory implies that all eigenvalues of $A$ will have negative real part only if $\sigma^2 N < 1$. This places a sharp constraint on both diversity $N$ and interaction strengh $\sigma$, often referred to as "complexity begets instability". (This argument generalizes to $\langle B_{ij}\rangle \neq 0$, to incomplete connectivity, or to correlated interactions.)

In ecology, many authors have sought to ease the tension between May's prediction and empirical observation by invoking effects not captured by dynamical systems with random coefficients: spatial or food-web structure, time delays, etc. Here, we show that May's argument is incomplete: random dynamical systems do not necessarily imply that stability decreases with diversity or interaction strength---the opposite behavior is also possible, without the need for special or additional structure.

@@

Consider the dynamical system in $N$ variables
\begin{equation}\label{dynamics}
    \dot{x}_i = f(x_i) + \sum_{j}a_{ij}g(x_i)h(x_j) \, .
\end{equation}
Here $f(x_i)$ represents the contribution of a population $i$ to its own growth (its "production function"), while $g(x_i)$ and $h(x_j)$ capture the interaction of $i$ with other populations. That interaction is weighted by a coefficient $a_{ij}$, such that $a_{ij} > 0$ (resp. $a_{ji} < 0$) implies a positive (resp. negative) effect of $j$ on the growth of $i$. In the following we call $g(x_i)$ the "response function".

The classic generalized Lotka-Volterra competition model corresponds to $f$, $g$, $h$ all linear. However, it is natural both physically and biologically to consider more general functions, including power laws $f(x)\sim x^\alpha$, $g(x)\sim x^\beta$, $h(x) \sim x^\gamma$. From a physical perspective, we can imagine populations $x_i$ forming three-dimensional clusters whose growth is limited to their two-dimensional surface, leading to a production function $f(x) \sim x^{2/3}$. Biologically, the growth of organisms (populations of cells) has long been known to scale like $f(x) \sim x^k$ with $k\simeq 3/4$, which can be understood in terms of hydrodynamic constraints on vascular and pulmonary networks. For reasons that are not currently understood, a similar pattern of growth appears to recur at the community level. It is also common to model predator-prey interactions with a square root laws ($g(x) \sim h(x) \sim x^{1/2}$), for example.

@@

Now we decompose the interaction matrix as
$a = \mu \mathbf{1} + (\mu_s-\mu)I + S$,
with $\mathbf{1}$ the rank-one matrix with all entries equal to $1$
and $S$ a random matrix as defined above.
We can limit ourselves to consider 
$a = (\mu_s-\mu)I + S$ by relying on 
rank-one perturbation theory, we would only neglect outliers
uninfluential for the linear stbaility properties of the system by doing so
(see Ref.~[Stone2018]).
The Jacobian in Eqs.~\eqref{eq: jac off-diag} and \eqref{eq: jac diag} assumes therefore the form $M + LJR$ with
\begin{align}
    M & = \textrm{diag}\left(f_*'(\mathbf x) -
    \frac{g_*'(\mathbf x)f_*(\mathbf x)}{g_*(\mathbf x)}
    +(\mu_s-\mu)g_*(\mathbf x)h_*'(\mathbf x)\right) \ , \\
    L &= \textrm{diag}(g_*(\mathbf x)) \ , \\
    R &= \textrm{diag}(h_*'(\mathbf x)) \ .
\end{align}
The domain of the eigenvalues of $J^*$, defined by
Eq.~\eqref{eq: domain}, 
first touches the right half-plane at $\zeta = 0$, 
hence linear stability requires   
\begin{equation}\label{eq: random-stability}
    \sum_{i=1}^N \cfrac{\left(\cfrac{g_*(x_i)h_*'(x_i)}{f_*(x_i)}\right)^2}{
        \Bigl |\left(\cfrac{f_*'(x_i)}{f_*(x_i)} -
        \cfrac{g_*'(x_i)}{g_*(x_i)}\right)
        +(\mu_s-\mu) \cfrac{g_*(x_i)h_*'(x_i)}{f_*(x_i)} \Bigl |^2}
    < \sigma^{-2}. 
\end{equation}
This expression can be considered the
random interaction case equivalent of
the expression in Eq.~\eqref{homogeneous-general} 
which appears again in the denominator of the summand.
It reduces indeed to the homogenous case for $\sigma\to 0$.
\red{To be discussed.}

@@

In the general case in which the dependence on the equilibria
$\mathbf{x}_*$ does not cancel out, is not possible to
obtain a clean complexity-stability relationship as before.
We can make progress in the large $N$ limit by
trasforming the sum in an integral 
$\sum_{i=1}^N\to N\int dx_*P(x_*)$, where $P(x_*)$ 
is the equilibrium distribution of the system
depending on the statistic of the interactions,
and write
\begin{equation}\label{eq: random-stability int}
    \int dx_* \cfrac{NP(x_*) \left(\cfrac{g_*(x)h_*'(x)}{f_*(x)}\right)^2}{
        \Bigl |\left(\cfrac{f_*'(x)}{f_*(x)} -
        \cfrac{g_*'(x)}{g_*(x)}\right)
        +(\mu_s-\mu) \cfrac{g_*(x)h_*'(x)}{f_*(x)} \Bigl |^2}
    < \sigma^{-2} \ .
\end{equation}
\red{There are things to clarify and discuss.}
This expression allows us to obtain insights on the stability properties of
the system in the case in which we are able to compute or estimate 
the probability distribution function $P(x_*)$.

@@

The equilibrium probability distribution function for $x_*$, $P(x_*)$,
can be obtained as the pushforward of the distribition of $\xi$.
The change of measure
\begin{equation}
    \left|\frac{d \xi}{d x_*}\right|=\frac{|\alpha-\beta|x_*^{\alpha-\beta-1}}{\sigma \sqrt{N\langle x_*^{2\gamma}\rangle}} \, ,
\end{equation}
leads to
\begin{equation}\label{eq: dist general}
    P(x_*)=\frac{|\alpha-\beta|x_*^{\alpha-\beta-1}}{\sqrt{2\pi\sigma^2 N\langle x_*^{2\gamma}\rangle}}
    \exp{\left\{-\frac{(x_*^{\alpha-\beta}-\mu N\langle x_*^{\gamma}\rangle)^2}{2\sigma^2N\langle x_*^{2\gamma}\rangle}\right\}} \, ,
\end{equation}
where the expectations have to be computed self-consistently, 
taking also in account normalization constraint.

Let us focus to the case $\alpha=1$, $\beta=2$ and $\gamma=1$, with
$a_s=0$ in order to leverage the equilibrium solution in Eq.~\eqref{eq: dist general}.

@@

\begin{itemize}
    \item Equilibrium distribution and histogram for "anti-May" example
    \item ($\mu N$, $\sigma\sqrt{N}$) plane for "May" and "anti-May" examples
        with stability lines.
    \item ($\alpha$, $\beta$) plane with homogeneous interactions line.
    \item Others?
\end{itemize}
